Benchmarking Parameterized Quantum Circuit Learning for Quantum Chemical Applications

### 2025-03-04 manuscript/main.tex
validate on real device + figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{A Combinatorial Search of Parameterized Quantum Circuit Learning for Quantum Chemical Applications}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Grier M. Jones}
\affiliation[UTSG ECE]{
The Edward S. Rogers Sr. Department of Electrical and Computer Engineering, 
University of Toronto, 
10 Kings College Road, Toronto, Ontario, 
Canada M5S 3G4}
\alsoaffiliation[UTM CHEM]{
Department of Chemical and Physical Sciences, 
University of Toronto Mississauga, 
3359 Mississauga Road, Mississauga, Ontario, 
Canada L5L 1C6}

@@

[Problem: QML methods related to quantum chemistry are rarely studied so we chose two different sets of data, one that's classical in-quantum out and the other that is quantum in-quantum out]

[FOCUS ON QUANTUM CHEMISTRY]

@@

Similar to the aforementioned machine learning methods, quantum computing (QC) for chemical applications has become increasingly popular due to possible quantum speedups for quantum chemical calculations.\cite{cao_quantum_2019}
While most quantum computing studies focus on creating new and more efficient electronic structure methods on quantum computers\cite{romero_strategies_2019,mcardle_quantum_2020,bauer_quantum_2020,cao_quantum_2019}, an approach that combines both ML and QC, is quantum machine learning (QML).
In recent years, QML has shown potential for quantum speedups, either through formal mathematical proofs or numerical experiments based on empirical observations.\cite{biamonte_quantum_2017}
Among the classes of QML algorithms, parameterized quantum circuits (PQCs), which are based on a hybrid quantum--classical optimization loop, have shown promise as flexible QML models, capable of achieving non-trivial results on near-term quantum hardware.\cite{benedetti_parameterized_2019,suzuki_predicting_2020}
In the context of chemistry, quantum machine learning (QML) has gained traction for tasks such as drug discovery\cite{suzuki_predicting_2020,smaldone_quantum--classical_2024,bhatia_quantum_2023,kao_exploring_2023,li_quantum_2021,avramouli_quantum_2023,avramouli_unlocking_2023}, materials discovery\cite{ishiyama_noise-robust_2022,ryu_quantum_2023,vitz_hybrid_2024}, proton affinities\cite{jin_integrating_2025}, and experimental molecular properties, including the log solubility in water, melting point, octanol/water distribution coefficient, hydration free energy of small molecules in water.\cite{hatakeyama-sato_quantum_2023}

@@

\section{Methods}
PQCs are often constructed of three parts: encoding layers that are used to encode the features onto a quantum circuit, variational layers which include parameters that are optimized classically, and measurements which provide numerical estimations of the regression target values.\citep{suzuki_predicting_2020} 
In this study, we utilize the Mitarai (M)\cite{mitarai_quantum_2018}, single- (A1) and double-angle (A2) encoding layers found in Ref. \citep{suzuki_predicting_2020}, along with the instantaneous quantum polynomial (IQP) circuit found in Refs. \citep{bremner_average-case_2016} and \citep{havlicek_supervised_2019}.
In the following section, we follow the notations derived from  Ref. \citep{suzuki_predicting_2020}.
Encoding layers work mapping a $d$-dimensional feature vector, $\mathbf{x}=(x_{1}, x_{2}, \ldots, x_{d})^{T} \in \mathbb{R}^{d}$, normalized on the range $[-1,1]$, onto a quantum circuit using a unitary matrix, denoted as $U_{\Phi(\mathbf{x})}$, to produce the quantum state $U_{\Phi(\mathbf{x})}\ket{0}^{\otimes n}$, where $n$ are the number of qubits.
The encoding layer takes the following general form,
\begin{equation}
	U_{\Phi(x)} =  \prod_{l} E_{\text{ent}}^{l} U_{\phi_{l}(\mathbf{x})}
	\label{eq:general_encoding}
\end{equation}
where, $E_{\text{ent}}^{l}$ denotes the entangling gates, which can be a CNOT, CZ, or identity ($\mathbf{I}$) gates,   $U_{\phi_{l}(\mathbf{x})}$ denotes the choice of encoding unitaries. 
Like in  Ref. \citep{suzuki_predicting_2020}, we choose $l \in \{1, 2\}$, such that when $l=1$, $E_{\text{ent}}$ corresponds to an identity matrix and $U_{\Phi(x)} = U_{\phi_{1}(\mathbf{x})}$.

@@

Following the encoding layers,variational (or ansatz) layers are used to introduce trainable parameters into the quantum circuit.
We use a mixed notation from Refs. \citep{suzuki_predicting_2020} and \citep{sim_expressibility_2019}, since Ref. \citep{sim_expressibility_2019} contains all of the variational layers used within this work.
We relegate the discussion of the expressibility and entanglement examined in that work to Section \ref{section:results_and_discussion}.
A general variational layer can be denoted as,
\begin{equation}
	U(\bm{\theta}) = \prod_{v} U_{v}(\bm{\theta}_{v}), % E_{\text{ent}}^{v}
	\label{eq:general_variational}
\end{equation}
where $\bm{\theta}$ denotes the variational parameters and $v$ denotes the number of times that the layer is repeated within the circuit. 
As $v$ increases and the number of trainable parameters ($\bm{\theta}$) increase, the theoretical assumption is that the model expressibility should also increase.
In our study, we choose $v \in \{1, 3, 5\}$ and refer to this as the number of ansatz layers (ALs).
We examine 12 different variational circuits, as shown in Fig. \ref{fig:ansatz}, which are denoted using the following labels: Modified-Pauli-CRZ (Fig. \ref{fig:ansatz}\textbf{(a)}) , Modified-Pauli-CRX (Fig. \ref{fig:ansatz}\textbf{(b)}), Efficient-CRZ (Fig. \ref{fig:ansatz}\textbf{(c)}), Efficient-CRX (Fig. \ref{fig:ansatz}\textbf{(d)}),, HWE-CNOT (Fig. \ref{fig:ansatz}\textbf{(e)}), HWE-CZ (Fig. \ref{fig:ansatz}\textbf{(f)}), ESU2 (Fig. \ref{fig:ansatz}\textbf{(g)}), Full-Pauli-CRZ (Fig. \ref{fig:ansatz}\textbf{(h)}), Full-Pauli-CRX (Fig. \ref{fig:ansatz}\textbf{(i)}), Hadamard (Fig. \ref{fig:ansatz}\textbf{(j)}), Full-CRZ (Fig. \ref{fig:ansatz}\textbf{(k)}), and Full-CRX (Fig. \ref{fig:ansatz}\textbf{(l)}),

@@

Used QisKit for Fake backend

@@

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Benchmarking Parameterized Quantum Circuit Learning for Quantum Chemical Applications}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Grier M. Jones}
\affiliation[UTSG ECE]{
The Edward S. Rogers Sr. Department of Electrical and Computer Engineering, 
University of Toronto, 
10 Kings College Road, Toronto, Ontario, 
Canada M5S 3G4}
\alsoaffiliation[UTM CHEM]{
Department of Chemical and Physical Sciences, 
University of Toronto Mississauga, 
3359 Mississauga Road, Mississauga, Ontario, 
Canada L5L 1C6}

@@

[General problem and classical machine learning]

@@

[quantum computing and qml]

Similar to the aforementioned machine learning methods, quantum computing (QC) for chemical applications has become increasingly popular due to possible quantum speedups for quantum chemical calculations.\cite{cao_quantum_2019}
While most quantum computing studies focus on creating new and more efficient electronic structure methods on quantum computers\cite{romero_strategies_2019,mcardle_quantum_2020,bauer_quantum_2020,cao_quantum_2019}, an approach that combines both ML and QC, is quantum machine learning (QML).
In recent years, QML has shown potential for quantum speedups, either through formal mathematical proofs or numerical experiments based on empirical observations.\cite{biamonte_quantum_2017}

[Parameterized quantum circuits]

Among the classes of QML algorithms, parameterized quantum circuits (PQCs), which are based on a hybrid quantum--classical optimization loop, have shown promise as flexible QML models, capable of achieving non-trivial results on near-term quantum hardware.\cite{benedetti_parameterized_2019,suzuki_predicting_2020}
In the context of chemistry, quantum machine learning (QML) has gained traction for tasks such as drug discovery\cite{suzuki_predicting_2020,smaldone_quantum--classical_2024,bhatia_quantum_2023,kao_exploring_2023,li_quantum_2021,avramouli_quantum_2023,avramouli_unlocking_2023}, materials discovery\cite{ishiyama_noise-robust_2022,ryu_quantum_2023,vitz_hybrid_2024}, proton affinities\cite{jin_integrating_2025}, and experimental molecular properties, including the log solubility in water, melting point, octanol/water distribution coefficient, hydration free energy of small molecules in water.\cite{hatakeyama-sato_quantum_2023}

[What we introduce; Problem: QML methods related to quantum chemistry are rarely studied so we chose two different sets of data, one that's classical in-quantum out and the other that is quantum in-quantum out; FOCUS ON QUANTUM CHEMISTRY]

@@

\section{Methods}
PQCs are often constructed of three parts: encoding layers that are used to encode the features onto a quantum circuit, variational layers which include parameters that are optimized classically, and measurements which provide numerical estimations of the regression target values.\citep{suzuki_predicting_2020} 
In this study, we utilize the Mitarai (M)\cite{mitarai_quantum_2018}, single- (A1) and double-angle (A2) encoding layers found in Ref. \citep{suzuki_predicting_2020}, along with the instantaneous quantum polynomial (IQP) circuit found in Refs. \citep{bremner_average-case_2016} and \citep{havlicek_supervised_2019}.
In the following section, we follow the notations derived from  Ref. \citep{suzuki_predicting_2020}.
Encoding layers work mapping a $d$-dimensional feature vector, $\mathbf{x}=(x_{1}, x_{2}, \ldots, x_{d})^{T} \in \mathbb{R}^{d}$, normalized on the range $[-1,1]$, onto a quantum circuit using a unitary matrix, denoted as $U_{\Phi(\mathbf{x})}$, to produce the quantum state $U_{\Phi(\mathbf{x})}\ket{0}^{\otimes n}$, where $n$ are the number of qubits.
The encoding layer takes the following general form,
\begin{equation}
	U_{\Phi(x)} =  \prod_{l} E_{\text{ent}}^{l} U_{\phi_{l}(\mathbf{x})}
	\label{eq:general_encoding}
\end{equation}
where, $E_{\text{ent}}^{l}$ denotes the entangling gates, which can be a CNOT, CZ, or identity ($\mathbf{I}$) gates,   $U_{\phi_{l}(\mathbf{x})}$ denotes the choice of encoding unitaries. 
Like in  Ref. \citep{suzuki_predicting_2020}, $E_{\text{ent}}$ corresponds to an identity matrix and $U_{\Phi(x)} = U_{\phi_{1}(\mathbf{x})}$.

%Like in  Ref. \citep{suzuki_predicting_2020}, we choose $l \in \{1, 2\}$, such that when $l=1$, $E_{\text{ent}}$ corresponds to an identity matrix and $U_{\Phi(x)} = U_{\phi_{1}(\mathbf{x})}$.

@@

Following the encoding layers,variational (or ansatz) layers are used to introduce trainable parameters into the quantum circuit.
We use a mixed notation from Refs. \citep{suzuki_predicting_2020} and \citep{sim_expressibility_2019}, since Ref. \citep{sim_expressibility_2019} contains all of the variational layers used within this work.
We relegate the discussion of the expressibility and entanglement examined in that work to Section \ref{section:results_and_discussion}.
A general variational layer can be denoted as,
\begin{equation}
	U(\bm{\theta}) = \prod_{v} U_{v}(\bm{\theta}_{v}), % E_{\text{ent}}^{v}
	\label{eq:general_variational}
\end{equation}
where $\bm{\theta}$ denotes the variational parameters and $v$ denotes the number of times that the layer is repeated within the circuit. 
As $v$ increases and the number of trainable parameters ($\bm{\theta}$) increase, the theoretical assumption is that the model expressibility should also increase.
In our study, we the number of variational layers as the number of ansatz layers (ALs).
%In our study, we choose $v \in \{1, 3, 5\}$ and refer to this as the number of ansatz layers (ALs).
We examine 12 different variational circuits, as shown in Fig. \ref{fig:ansatz}, which are denoted using the following labels: Modified-Pauli-CRZ (Fig. \ref{fig:ansatz}\textbf{(a)}) , Modified-Pauli-CRX (Fig. \ref{fig:ansatz}\textbf{(b)}), Efficient-CRZ (Fig. \ref{fig:ansatz}\textbf{(c)}), Efficient-CRX (Fig. \ref{fig:ansatz}\textbf{(d)}),, HWE-CNOT (Fig. \ref{fig:ansatz}\textbf{(e)}), HWE-CZ (Fig. \ref{fig:ansatz}\textbf{(f)}), ESU2 (Fig. \ref{fig:ansatz}\textbf{(g)}), Full-Pauli-CRZ (Fig. \ref{fig:ansatz}\textbf{(h)}), Full-Pauli-CRX (Fig. \ref{fig:ansatz}\textbf{(i)}), Hadamard (Fig. \ref{fig:ansatz}\textbf{(j)}), Full-CRZ (Fig. \ref{fig:ansatz}\textbf{(k)}), and Full-CRX (Fig. \ref{fig:ansatz}\textbf{(l)}),

@@

Used QisKit for real and fake back end using Qiskit Batches

optimization levels
none (0)

``No optimization: typically used for hardware characterization
Basic translation and Layout/Routing: TrivialLayout, where it selects the same physical qubit numbers as virtual and inserts SWAPs to make it work (using StochasticSwap)''

light (1)
``Light optimization: Layout/Routing: Layout is first attempted with TrivialLayout. If additional SWAPs are required, a layout with a minimum number of SWAPs is found by using SabreSWAP, then it uses VF2LayoutPostLayout to try to select the best qubits in the graph, InverseCancellation, 1Q gate optimization''

medium (2)
``Medium optimization: Layout/Routing: Optimization level 1 (without trivial) + heuristic optimized with greater search depth and trials of optimization function. Because TrivialLayout is not used, there is no attempt to use the same physical and virtual qubit numbers. CommutativeCancellation''

high (3)
``High Optimization:
Optimization level 2 + heuristic optimized on layout/routing further with greater effort/trials
Resynthesis of two-qubit blocks using Cartan's KAK Decomposition.
Unitarity-breaking passes:
OptimizeSwapBeforeMeasure: Moves the measurements around to avoid SWAPs
RemoveDiagonalGatesBeforeMeasure: Removes gates before measurements that would not effect the measurements''

resilience levels (error mitigation)
none (0) 
1 ``Minimal mitigation costs: Mitigate error associated with readout errors'' ``Twirled Readout Error eXtinction (TREX) measurement twirling''
Medium mitigation cost medium 
2 ``Medium mitigation costs. Typically reduces bias in estimators, but is not guaranteed to be zero-bias.'' ``Level 1 + Zero Noise Extrapolation (ZNE) and gate twirling''

@@

Talk about the cost of going wider, and inherently deeper

@@

Ran using the state vector model parameters for one iteration to test the optimization and resilience levels using Fake Quebec before running on the real device

$\{1024 \times x \vert  x \in [1,10]\}$

Optimization level 2 
resilience level 0

Regarding the number of circuit executions vs performance 3072 (1024 times 3) is the best number of shots

Fig. \ref{fig:optresfakequebec}
Fig. \ref{fig:shotsfakequebec}
